---
title: "Severe Weather Events Visualized"
author: "Zack Baker"
date: "01/06/2020"
output: 
  html_document:
   theme: paper
   highlight: textmate
   toc: true
   toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Synopsis

NOAA (National Oceanic & Atmospheric Administration) is the world's largest provider of weather and climate data. In particular with severe weather events such as blizzards and tornadoes, the [Storm Events Database](https://www.ncdc.noaa.gov/stormevents/) collects information on significant weather phenomena, dating back to January 1950. The following report aims to answer questions on the types of events that have caused the greatest health and economic impact since such time. 


### Data Processing

Note: From Jan 1950 to Dec 1995, only Tornado, Thunderstorm Wind and Hail had been recorded as Severe weather events. As such, to depict a true picture of impact for all 48 event types, the dataset will be from 1996 to 2011.



##### Required R Libraries

```{r results="hide", warning=FALSE, message=FALSE}
library(data.table)
library(stringr)
library(dplyr)

```


##### Read Storm Data and Process  

```{r results="hide", cache=FALSE}

csv_name <- "repdata_data_StormData.csv.bz2"

# Read file
storm_data <-fread(csv_name)

# Format date variables
storm_data <- storm_data[,BDate := as.POSIXct(word(BGN_DATE,1), format = "%m/%d/%Y")]

# Reduce dataset to 1996 onwards
storm_data <- storm_data[format(BDate, "%Y")> 1995,]

# Tally Damage from property and crops
storm_data$TotalDamage <- 
         ifelse(storm_data$PROPDMGEXP=="M", storm_data$PROPDMG*1000000, ifelse(storm_data$PROPDMGEXP=="K", storm_data$PROPDMG*1000,0)) +
         ifelse(storm_data$CROPDMGEXP=="M", storm_data$CROPDMG*1000000, ifelse(storm_data$CROPDMGEXP=="K", storm_data$CROPDMG*1000,0))

```




### Storm Data Questions  

#### Question 1
1) Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

```{r}

# Group By Fatalities
data_agg_deaths <- storm_data %>% 
mutate(Year = format(BDate, "%Y")) %>% group_by(EVTYPE) %>% summarise(total = sum(FATALITIES))

# Arrange and Remove zero totals
data_agg_deaths <- data_agg_deaths[data_agg_deaths$total > 0,]
data_agg_deaths <- data_agg_deaths %>% arrange(desc(total))

# EVtypes from top 5% of total deaths between 1996-2011
top_data_agg_deaths <- data_agg_deaths[data_agg_deaths$total > quantile(data_agg_deaths$total, prob=1-5/100),]

colnames(top_data_agg_deaths) <- c("EventType", "Total")

# Group By Injuries
data_agg_injured <- storm_data %>% 
mutate(Year = format(BDate, "%Y")) %>% group_by(EVTYPE) %>% summarise(total = sum(INJURIES))

# Arrange and Remove zero totals
data_agg_injured <- data_agg_injured[data_agg_injured$total > 0,]
data_agg_injured <- data_agg_injured %>% arrange(desc(total))

# EVtypes from top 5% of total injuries between 1996-2011
top_data_agg_injured <- data_agg_injured[data_agg_injured$total > quantile(data_agg_injured$total, prob=1-5/100),]

colnames(top_data_agg_injured) <- c("EventType", "Total")

```


##### Plot Fatalities by EV Types

```{r }


barplot(table(top_data_agg_deaths$EventType), height=top_data_agg_deaths$Total, names=top_data_agg_deaths$EventType, col=rainbow(length(top_data_agg_deaths$EventType)), cex.name=0.5, cex.axis = 0.75, las=2, horiz=T, main="U.S. Fatalities by Event Type Between 1996-2011")



```


##### Plot Injuries by EV Types

```{r}


barplot(table(top_data_agg_injured$EventType), height=top_data_agg_injured$Total, names=top_data_agg_injured$EventType, col=rainbow(length(top_data_agg_injured$EventType)), cex.name=0.5, cex.axis = 0.75, las=2, horiz=T, main="U.S. Injuries by Event Type Between 1996-2011")



```


#### Question 2
2) Across the United States, which types of events have the greatest economic consequences?  


```{r}



# Group By Damages
data_agg_damages <- storm_data %>% 
mutate(Year = format(BDate, "%Y")) %>% group_by(EVTYPE) %>% summarise(total = sum(TotalDamage))

# Arrange and Remove zero totals
data_agg_damages <- data_agg_damages[data_agg_damages$total > 0,]
data_agg_damages <- data_agg_damages %>% arrange(desc(total))

# EVtypes from top 5% of total damages between 1996-2011
top_data_agg_damages <- data_agg_damages[data_agg_damages$total > quantile(data_agg_damages$total, prob=1-5/100),]


top_data_agg_damages$total <- top_data_agg_damages$total/1000000000

top_data_agg_damages$total <- format(round(top_data_agg_damages$total,2), nsmall=2)

top_data_agg_damages$total <- as.numeric(top_data_agg_damages$total)

colnames(top_data_agg_damages) <- c("EventType", "USD_BILLIONS")

```


##### Plot Economic Impact ($)

```{r}

barplot(table(top_data_agg_damages$EventType), height=top_data_agg_damages$USD_BILLIONS, names=top_data_agg_damages$EventType, col=rainbow(length(top_data_agg_damages$EventType)), cex.name=0.5, cex.axis = 0.75, las=2, horiz=T, main="Economic Impact by Event Type Between 1996-2011 (In USD Billions)")

```

