---
title: "NOAAData"
author: "Zack"
date: "01/06/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Loading Data  

```{r results="hide", warning=FALSE, message=FALSE}
library(data.table)
library(stringr)
library(dplyr)
library(ggplot2)

```


Read Storm Data and Process  

```{r results="hide", cache=FALSE}
csv_name <- "repdata_data_StormData.csv.bz2"

# Read file
storm_data <-fread(csv_name)

# Format date and time variables
storm_data <- storm_data[,BDate := as.POSIXct(word(BGN_DATE,1), format = "%m/%d/%Y")]
storm_data <- storm_data[,BTime := paste0(str_sub(as.character(BGN_TIME),1,2),":", str_sub(as.character(BGN_TIME),-2,-1))]

storm_data$TotalDamage <- 
         ifelse(storm_data$PROPDMGEXP=="M", storm_data$PROPDMG*1000000, ifelse(storm_data$PROPDMGEXP=="K", storm_data$PROPDMG*1000,0)) +
         ifelse(storm_data$CROPDMGEXP=="M", storm_data$CROPDMG*1000000, ifelse(storm_data$CROPDMGEXP=="K", storm_data$CROPDMG*1000,0))



```


##### Storm Data Questions  


1) Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

```{r cache=FALSE}

# Group By Fatalities
data_agg_deaths <- storm_data %>% 
mutate(Year = format(BDate, "%Y")) %>% group_by(EVTYPE) %>% summarise(total = sum(FATALITIES))

# Arrange and Remove zero totals
data_agg_deaths <- data_agg_deaths[data_agg_deaths$total > 0,]
data_agg_deaths <- data_agg_deaths %>% arrange(desc(total))

# EVtypes from top 5% of total deaths between 1950-2011
top_data_agg_deaths <- data_agg_deaths[data_agg_deaths$total > quantile(data_agg_deaths$total, prob=1-5/100),]

colnames(top_data_agg_deaths) <- c("EventType", "Total")

# Group By Injuries
data_agg_injured <- storm_data %>% 
mutate(Year = format(BDate, "%Y")) %>% group_by(EVTYPE) %>% summarise(total = sum(INJURIES))

# Arrange and Remove zero totals
data_agg_injured <- data_agg_injured[data_agg_injured$total > 0,]
data_agg_injured <- data_agg_injured %>% arrange(desc(total))

# EVtypes from top 5% of total injuries between 1950-2011
top_data_agg_injured <- data_agg_injured[data_agg_injured$total > quantile(data_agg_injured$total, prob=1-5/100),]

colnames(top_data_agg_injured) <- c("EventType", "Total")

```


Plot Fatalities by EV Types

```{r}


barplot(table(top_data_agg_deaths$EventType), height=top_data_agg_deaths$Total, names=top_data_agg_deaths$EventType, col=rainbow(length(top_data_agg_deaths$EventType)), cex.name=0.5, cex.axis = 0.75, las=2, horiz=T, main="U.S. Fatalities by Event Type Between 1950-2011")



```


Plot Injuries by EV Types

```{r}


barplot(table(top_data_agg_injured$EventType), height=top_data_agg_injured$Total, names=top_data_agg_injured$EventType, col=rainbow(length(top_data_agg_injured$EventType)), cex.name=0.5, cex.axis = 0.75, las=2, horiz=T, main="U.S. Injuries by Event Type Between 1950-2011")



```


2) Across the United States, which types of events have the greatest economic consequences?  


```{r}



# Group By Damages
data_agg_damages <- storm_data %>% 
mutate(Year = format(BDate, "%Y")) %>% group_by(EVTYPE) %>% summarise(total = sum(TotalDamage))

# Arrange and Remove zero totals
data_agg_damages <- data_agg_damages[data_agg_damages$total > 0,]
data_agg_damages <- data_agg_damages %>% arrange(desc(total))

# EVtypes from top 5% of total damages between 1950-2011
top_data_agg_damages <- data_agg_damages[data_agg_damages$total > quantile(data_agg_damages$total, prob=1-5/100),]


top_data_agg_damages$total <- top_data_agg_damages$total/1000000000

top_data_agg_damages$total <- format(round(top_data_agg_damages$total,2), nsmall=2)

top_data_agg_damages$total <- as.numeric(top_data_agg_damages$total)

colnames(top_data_agg_damages) <- c("EventType", "USD_BILLIONS")

```


Plot Economic Impact ($)

```{r}

barplot(table(top_data_agg_damages$EventType), height=top_data_agg_damages$USD_BILLIONS, names=top_data_agg_damages$EventType, col=rainbow(length(top_data_agg_damages$EventType)), cex.name=0.5, cex.axis = 0.75, las=2, horiz=T, main="Economic Impact by Event Type Between 1950-2011 (In USD Billions)")

```

